---
title: "Selection Simulation: Exploratory Dashboard"
subtitle: "Allele fixation and inbreeding depression under long-term selection"
format:
  html:
    toc: true
    theme: cosmo
    embed-resources: true
bibliography: ../references.bib
---

```{r setup}
#| include: false
library(arrow)
library(duckdb)
library(DBI)
library(dplyr)
library(ggplot2)
library(knitr) # not called directly below, but Quarto's R-chunk execution engine needs it
                # installed; renv's implicit dependency scan only picks up an explicit library()
                # call, not Quarto's own internal use of it (see R/io.R for the same pattern with
                # dplyr and arrow::write_dataset()).

data_dir <- Sys.getenv("SELECTION_DATA_DIR", unset = "../data")
con <- dbConnect(duckdb::duckdb())

pal <- c(low = "#2a78d6", high = "#eb6834")
theme_dashboard <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e1e0d9"),
    axis.line = element_line(color = "#c3c2b7"),
    axis.text = element_text(color = "#898781"),
    legend.position = "top"
  )
```

## Introduction

This dashboard reads directly from the Parquet lake produced by `run.R`, queried through
DuckDB, and shows what the current run's `data/` directory contains: additive genetic variance
over time and allele fixation progress, both reported as distributions across replicates rather
than single trajectories, since a lone run of a stochastic process is close to uninformative on
its own. The underlying engine is AlphaSimR [@gaynor2021alphasimr], run over long horizons with
recurrent mutation, because a fixed founder haplotype set exhausts standing variation and
selection response flatlines as a modeling artifact rather than a biological one
[@robertson1960limits; @hill1986recurrent].

## Selection response across replicates

```{r va-query}
va_by_rep <- dbGetQuery(con, sprintf(
  "SELECT scenario_id, replicate_id, generation, va
   FROM read_parquet('%s/summary_metrics/**/*.parquet', hive_partitioning=1)
   ORDER BY scenario_id, replicate_id, generation",
  data_dir
))

va_by_rep$intensity <- sub("^.*_(low|high)$", "\\1", va_by_rep$scenario_id)

va_bands <- va_by_rep |>
  group_by(intensity, generation) |>
  summarise(mean_va = mean(va), min_va = min(va), max_va = max(va), .groups = "drop")
```

```{r va-plot}
#| fig-cap: "Mean additive genetic variance per generation (line), with the min-max range across replicates shown as a band. Each replicate carries its own independently generated founder population and RNG stream (see Limitations)."
ggplot(va_bands, aes(x = generation, color = intensity, fill = intensity)) +
  geom_ribbon(aes(ymin = min_va, ymax = max_va), alpha = 0.2, color = NA) +
  geom_line(aes(y = mean_va), linewidth = 0.8) +
  scale_color_manual(values = pal, name = "Selection intensity") +
  scale_fill_manual(values = pal, name = "Selection intensity") +
  labs(x = "Generation", y = "Additive genetic variance (mean, min-max band)") +
  theme_dashboard
```

Variance depletes fastest under high selection intensity, as expected: fewer parents contribute
to each generation, so both selection and drift consume standing variation more quickly. Recurrent
mutation, calibrated from a target mutational heritability rather than a molecular per-bp rate
[@houle1996mutational; @lynch1998qtl], keeps a non-trivial trickle of new variance flowing in
rather than letting the population run to complete fixation.

## Fixation progress across replicates

```{r fixation-query}
seg_by_rep <- dbGetQuery(con, sprintf(
  "SELECT scenario_id, replicate_id, generation, COUNT(DISTINCT site_id) AS n_segregating
   FROM read_parquet('%s/per_locus_freq/**/*.parquet', hive_partitioning=1)
   GROUP BY scenario_id, replicate_id, generation
   ORDER BY scenario_id, replicate_id, generation",
  data_dir
))
seg_by_rep$intensity <- sub("^.*_(low|high)$", "\\1", seg_by_rep$scenario_id)

seg_bands <- seg_by_rep |>
  group_by(intensity, generation) |>
  summarise(mean_n = mean(n_segregating), min_n = min(n_segregating),
            max_n = max(n_segregating), .groups = "drop")
```

```{r fixation-plot}
#| fig-cap: "Count of still-segregating QTL sites per generation (mean line, min-max band across replicates). A falling count reflects loci reaching fixation; sites already fixed are dropped from the sampled table by design (segregating_only in the scenario config), so this tracks remaining standing variation directly rather than through variance alone."
ggplot(seg_bands, aes(x = generation, color = intensity, fill = intensity)) +
  geom_ribbon(aes(ymin = min_n, ymax = max_n), alpha = 0.2, color = NA) +
  geom_line(aes(y = mean_n), linewidth = 0.8) +
  scale_color_manual(values = pal, name = "Selection intensity") +
  scale_fill_manual(values = pal, name = "Selection intensity") +
  labs(x = "Generation", y = "Segregating QTL sites (mean, min-max band)") +
  theme_dashboard
```

```{r cleanup}
#| include: false
dbDisconnect(con, shutdown = TRUE)
```

## Validation

The four checks below run as a separate suite (`validation/`), not as part of this dashboard's
own render: re-running a multi-replicate validation suite on every dashboard build would be slow
and would duplicate work already done. What follows is not a pass/fail summary; it is the actual
output each test produced on its most recent run, reported in full alongside the production
metrics above so the proof sits next to the result it backs up. See `METHODS.md` for the
derivations and reasoning behind these numbers.

### Neutral heterozygosity decay [@falconer1996qtg; @lynch1998qtl]

Log-linear decay fit to the $\mu = 0$ validation-marker panel, five replicates per condition,
nominal $N_e = 100$:

| Condition | Implied $N_e$ (mean) | Range across replicates |
|---|---|---|
| No selection | 104.0 | 89.4 to 125.0 |
| Low-intensity selection | 14.1 | 9.9 to 17.6 |

### Breeder's equation, $R = h^2 S$ [@falconer1996qtg]

Cumulative gap between predicted and actual response, and realized heritability ($V_A / V_P$)
against the nominal 0.300 the prediction used throughout:

| Generation | Cumulative predicted-vs-actual gap | Realized $h^2$ |
|---|---|---|
| 5 | 2.5% | 0.24 (early mean) |
| 10 | 20.6% | |
| 25 | 94.1% | |
| 50 | 217.4% | |
| 100 | 476.5% | 0.02 (late mean) |

### $F_{ROH}$ against pedigree $F$ [@curik2014roh; @keller2011roh; @kardos2015genomic]

Founder population of 40, 8 parents per generation, 40 generations of deliberately intense
inbreeding:

| Generation | Mean $F_{ROH}$ | Mean pedigree $F$ |
|---|---|---|
| 5 | 0.27 | 0.22 |
| 40 | 0.94 | 0.90 |

Final-generation correlation across runs: $r \approx 0.34$ to $0.37$. Mean
$|F_{ROH} - F_{\text{pedigree}}|$: 0.04 to 0.07.

### Inbreeding depression [@doekes2021meta]

Ten replicates pooled per condition (12,500 observations), regressing phenotype on pedigree $F$
with a replicate-by-generation cohort fixed effect to remove genetic drift's own correlation with
generation number:

| Architecture | Slope | SE | $p$ | Slope in phenotypic SD per unit $F$ |
|---|---|---|---|---|
| Purely additive | 0.472 | 0.372 | 0.205 | 0.22 (not significant) |
| Directional dominance (mean dominance degree 0.6) | -2.369 | 0.364 | < 0.0001 | -1.23 |

The non-significant additive slope is not a weak result to explain away; it is the expected null,
since a purely additive trait cannot show mean inbreeding depression by construction, and
confirming that null was as much the point of this test as the dominance result was. The
directional-dominance architecture used here is not the one the scenario behind this dashboard's
own charts was generated under; extending it to the main pipeline is a separate decision, not yet
made. The comparison to the Doekes et al. benchmark of roughly 0.13% decline per 1% rise in $F$ is
qualitative: this trait has an arbitrary mean-zero scale, not a natural biological unit, so a
percent-of-mean comparison is not well defined, and the standard-deviation framing above is
reported instead of a claim of numeric equivalence.

## Limitations

- **Founder populations are not reproducible from a seed.** AlphaSimR's compiled coalescent
  function is not deterministic even given identical literal seeds. Everything *after* founders
  exist (mutation, selection, crossing, checkpoint and resume) is reproducible; the founder
  population itself is not, and is persisted as an artifact per replicate specifically because of
  this. Reproducing a specific run's founders requires that file, not its seed.
- **Inbreeding depression requires a separate trait architecture.** The scenarios behind this
  dashboard use a purely additive trait, which cannot show mean inbreeding depression by
  construction. The depression validation test builds its own directional-dominance population
  (`SP$addTraitAD`) to demonstrate the mechanism; extending dominance to the main pipeline is a
  separate decision, not yet made.
- **The Doekes comparison is qualitative, not literal.** This trait has an arbitrary mean-zero
  scale, not a natural biological unit like birth weight, so a percent-of-mean comparison to the
  Doekes et al. benchmark [@doekes2021meta] is not well defined; the validation test reports a
  phenotypic-standard-deviation-based slope instead, which is directionally comparable but not a
  claim of numeric equivalence.
- **MoBPS remains the fallback** if AlphaSimR's founder non-determinism or its other constraints
  become limiting for the recurrent-mutation and OCS features this project may need later
  [@pook2020mobps].

## References
